A general and efficient representation of Ancestral Recombination Graphs

### 2023-08-08 paper.tex

% JK: this is a rough first pass for a slightly different paper. Needs
% substantial revision.
% https://www.overleaf.com/5594399339wckggzwhnfvv
\begin{abstract}
Recent breakthroughs have made it possible to infer genetic ancestry in
the presence of recombination at scale, enabling many
downstream applications in population and statistical genetics.
Such recombinant genetic ancestry is usually
referred to as an Ancestral Recombination Graph, or ARG.
% Note: slight repetition here with first sentence.
There are now multiple methods that can infer ARGs at a practical scale
and it is therefore vital that these methods can be systematically evaluated
and compared to determine their strengths and weaknesses.
Unfortunately confusion exists over terminology and there is
little agreement on shared standards for data interchange,
significantly hampering progress in the field.
This lack of clarity and agreement on basic issues such as what does and
does not constitute an ARG is partly attributable to the historical
development of the term.
Originally rigorously defined as
a stochastic process (the coalescent with recombination) together with its
representation, an ARG is now understood to refer to any
concrete realisation of a recombinant genetic ancestry.
We show that the standard description of an ARG
in terms of common ancestor and recombination events
(inherited from the original stochastic process definition) is
ambiguous,
% is it ambiguous? How? Can we simply delete this?
fundamentally limited and an unsuitable basis for data interchange.
We provide a simple alternative definition of an ARG in which a node
corresponds to one of an individual's monoploid genomes and an edge
defines genetic inheritance between two nodes, with annotations on
each edge specifying which regions of the genome have
been inherited via that route.
We show how this definition can be used to generalise the encoding to
cover both the standard and several simplified versions of ARGs, as well as
enabling efficient access to local trees along the genome. Additionally, we
define some classifications and metrics of ARG nodes. Finally, we relate
these generalisations to the ARGs produced by a number of recently
developed ancestral inference methods.
% jk-note: this ends a bit flat - we should say other stuff like we clarify
% how the ARG relates the pedigree, etc, etc. Let's go through again when
% we have a reasonable draft.
\end{abstract}

>>>>

% FIXME this is weak, but gives the basic outline
\begin{abstract}
Recent breakthroughs have made it possible to infer genetic genealogies in
the presence of recombination at scale, enabling many
downstream applications in population and statistical genetics.
The structure representing such recombinant genetic ancestry
is usually referred to as an Ancestral Recombination Graph (ARG),
although there is some confusion about the interpretation and little
agreement on the concrete details.
We propose a concrete definition of ARGs that is suitable for modern
datasets, in terms of genomes and their patterns of inheritance,
and contrast this with the classical event-based definition which
makes several limiting assumptions.
We show that this new definition enables fine-grained
levels of precision about ancestral processes to be represented,
and forms the basis of an efficient computational framework.
\end{abstract}
